Individual and Differential Harm in Redistricting

### 2022-03-29 paper/paper.Rmd
clarify nonconstant group IDs

To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a set of mutually exclusive groups of voters, so that for each voter $i$, we have $i\in G_k$ for some group $G_k$.
We will say that a plan $q$ is *fair* in comparison to a counterfactual $\tilde Q$ if $\E[H(q)\mid I\in G_k]$ is constant across groups $G_k\in\Groups$.

@@

In competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with $\E[\seats(\dem, V, Q)\mid Q=q]$ and thus $\Delta H(q)$ as well.
But in more lopsided states, $\beta(\half)$ can give unintuitive results, which, if not interpreted carefully, can lead practictioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.

@@

We sample 1,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
For each of these sampled plans and the two gerrymanders, we calculate the utility, harm, and fairness measures, as well as the expected number of Democratic seats, the efficiency gap, partisan bias, and the mean-median score.[^1]
The distribution of these measures is shown in Figure \ref{fig:nj-pairs}.

[^1]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature[@katz21].

@@

```{=tex}
\vspace{1in}
\appendix
\renewcommand\thefigure{\thesection.\arabic{figure}}
\setcounter{figure}{0}
```

# Computing Harm Measures from Precinct-level Election Data {#app:formulas}

@@

The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlike a generalized linear model.
Specifically, we define $\gamma:[0,1]\to\R$, $$
\gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
$$ which is the inverse CDF of a Laplace distribution with rate 2.[^2]
Then a party's vote share in precinct $k$ and election $v$, notated here as $s_{kv}$, is modeled as $$
\gamma(s_{kv}) = \gamma(\overline s_k) + \eps_v, \qquad \eps_v \iid \Norm(0, \sigma^2_{\mathrm{hist}}),
$$ where $\overline{s}_k$ is the party baseline, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
For congressional elections in the United States from 1974--2018, we have $\widehat\sigma^2_\mathrm{hist}=0.03723$, and so draw predictive deviations from $\mathrm{t}_{22}(0, \widehat\sigma^2_\mathrm{hist})$.

[^2]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.

>>>>

To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a set of mutually exclusive groups of voters, so that for each voter $i$, we have $i\in G_k$ for some group $G_k$.
In general we will not require the group identities to be fixed with respect to elections.[^1]
This is because some group identities, like partisan preference, may themselves change election-to-election.

[^1]: Formally, we have random indicator variables $\ind\{i\in G_k\}$ for all $i$ and $k$, subject to the constraint that $\sum_k \ind\{i\in G_k\}=1$ for all $i$.
    These indicator variables may be correlated with $V$ and/or $\tilde Q$, in general.

Once we have defined a set of groups, we say that a plan $q$ is *fair* in comparison to a counterfactual $\tilde Q$ if $\E[H(q)\mid I\in G_k]$ is constant across groups $G_k\in\Groups$.

@@

In competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with $\E[\seats(\dem, V, Q)\mid Q=q]$ and thus $\Delta H(q)$ as well.
But in more lopsided states, $\beta(\half)$ can give unintuitive results, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.

@@

We sample 1,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
For each of these sampled plans and the two gerrymanders, we calculate the utility, harm, and fairness measures, as well as the expected number of Democratic seats, the efficiency gap, partisan bias, and the mean-median score.[^2]
The distribution of these measures is shown in Figure \ref{fig:nj-pairs}.

[^2]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature [@katz21].

@@

```{=tex}
\vspace{1in}
\appendix
\renewcommand\thefigure{\thesection.\arabic{figure}}
\setcounter{figure}{0}
```
# Computing Harm Measures from Precinct-level Election Data {#app:formulas}

@@

The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlike a generalized linear model.
Specifically, we define $\gamma:[0,1]\to\R$, $$
\gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
$$ which is the inverse CDF of a Laplace distribution with rate 2.[^3]
Then a party's vote share in precinct $k$ and election $v$, notated here as $s_{kv}$, is modeled as $$
\gamma(s_{kv}) = \gamma(\overline s_k) + \eps_v, \qquad \eps_v \iid \Norm(0, \sigma^2_{\mathrm{hist}}),
$$ where $\overline{s}_k$ is the party baseline, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
For congressional elections in the United States from 1974--2018, we have $\widehat\sigma^2_\mathrm{hist}=0.03723$, and so draw predictive deviations from $\mathrm{t}_{22}(0, \widehat\sigma^2_\mathrm{hist})$.

[^3]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
